untitled

### 2016-05-07 model.tex
general edits

\begin{align}
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi - \frac{1}{\rho}\nabla
  P, \label{momeq}\\ 
  &\frac{D\epsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\epsilon 
  \tstop \nabla P \right),\label{dusteq}  
\end{align}
where $D/Dt \equiv \p_t + \bm{v}\cdot\nabla$ is the Lagragian
derivative; 
\begin{align}
  \rho \equiv \rhog + \rhod
\end{align}
is the total density with $\rhog$ and $\rhod$ being the gas and dust
density, respectively; 
\begin{align}
  \bm{v} \equiv \frac{\rhog\bm{v}_\mathrm{g} + 
    \rhod\bm{v}_\mathrm{d}}{\rho}
\end{align}
is the center-of-mass velocity with $\bm{v}_\mathrm{g}$ and
$\bm{v}_\mathrm{d}$ being the gas and dust velocities, respectively; 
and 
\begin{align}
  \epsilon \equiv \frac{\rhod}{\rho}  = \frac{\tepsilon}{1+\tepsilon} 
\end{align}
is the dust-fraction and $\tepsilon=\rhod/\rhog$ is the usual
dust-to-gas ratio. The gas pressure $P$ is given by the chosen 
equation of state described below. The external potential $\Phi$ is
that for a star of mass $M_*$ at the origin 
\begin{align}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}},
\end{align}
where $G$ is the gravitational constant.

@@

\begin{align}
\frac{DP}{Dt} &= - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2}
                + \mathcal{C},  \label{eff_energy} \\
\mathcal{C}&\equiv c_s^2 \nabla\cdot\left[\tstop\left(1 - \frac{P}{c_s^2\rho}\right)\nabla
  P\right]
\end{align}

Eq. \ref{eff_energy} is the energy equation for an ideal gas of adiabatic index
$\gamma=1$, i.e. isothermal evolution, but now with additional source 
terms due to the imposed temperature profile, as well as dust-gas
friction.

@@

\section{Equilibrium}
To obtain an axisymmetric steady state with $\rho(r,z)$ and 
$\bm{v}=r\Omega(r,z)\hat{\bm{\phi}}$ where $\Omega = v_\phi/r$, 
we need to solve 
\begin{align}
  r\Omega^2 &= \frac{\p \Phi}{\p r} + \frac{1}{\rho}\frac{\p P}{\p
    r},\label{steady_momr}\\
  0 & = \frac{\p\Phi}{\p z} + \frac{1}{\rho}\frac{\p P}{\p z},\label{steady_momz}\\
  0 & = \nabla\cdot\left(\epsilon\tstop\nabla P\right) \label{steady_dust}
\end{align}
for $\rho$, $\Omega$ and $\epsilon$ with $P=P(\epsilon,\rho)$ given by
the equation of state (Eq. \ref{eos}).

@@

\subsection{Vertical shear}
The mixture possess vertical shear. To see this, we eliminate $\Phi$
between Eq. \ref{steady_momr}---\ref{steady_momz} to 
obtain 
\begin{align}\label{vshear}
  r\frac{\p \Omega^2}{\p z} &= \frac{\p\ln{\rho}}{\p r}\frac{\p}{\p
    z}\left[c_s^2(1-\epsilon)\right] - \frac{\p\ln{\rho}}{\p z}
  \frac{\p}{\p r} \left[c_s^2(1-\epsilon)\right]\\  
  & = \frac{1}{\rho}\left(\frac{\p P}{\p r}\frac{\p s}{\p z} -\frac{\p
    P}{\p z}\frac{\p s}{\p r} \right),\notag
\end{align}
where
\begin{align}
   s \equiv \ln \frac{P}{\rho}
\end{align}
is a pseudo-entropy for the mixture.

@@

\begin{align}
  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0,\\
  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, 
\end{align}
with $s = \ln{P/\rho^\gamma}$ and 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)$ is the
square of the epicylic frequency. We assume $\kappa^2>0$. 
These criteria are applicable to any adiabatic fluid when $s$
is the usual definition of entropy. For the dusty problem, we may take 
$s=\ln{(1-\epsilon)}+\mathrm{const.}$ from the definition of $s$ and  
the equation of state (Eq. \ref{eos}) with uniform sound-speed.

%\begin{align}
%  s \equiv \ln \frac{P}{\rho} = \ln{(1-\epsilon)} 
%\end{align}
% The first equality is
%similar to the usual defintion for gas entropy (with unit adiabatic
%index). The second follows from the imposed equation of state,
%Eq. \ref{eos}, with a constant sound-speed.

The stability conditions are thus
\begin{align}\label{dusty_solberg}
  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\epsilon &> 0,  \\
  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\epsilon}{\p
    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \epsilon}{\p r} \right) & > 0. 
\end{align}

In accretion disks, $\rhog$, $\rhod$ and $\epsilon$ are expected to
decrease away from the mid-plane
\begin{align*}  
  \frac{\p\rhog}{\p |z|} < 0,
%, \quad \frac{\p \rhog}{\p r} < 0,   
\end{align*}
and similarly for $\rhod$, $\epsilon$, and the dust-to-gas ratio $\tepsilon$. 
We may thus evaluate the stability criteria at $z>0$ without loss of
generality. We insert the vertical shear profile,
Eq. \ref{vshear}, into Eq. \ref{dusty_solberg} and write the result 
in terms of $\tepsilon$, to find

\begin{align} 
  &1 + \frac{\tepsilon h_\mathrm{g}^2}{\left(1+
    \tepsilon\right)^2}\frac{\Omega^2}{\kappa^2}\left(\frac{\p\ln{\rhog}}{\p\ln{r}}
  \frac{\p\ln{\tepsilon}}{\p\ln{r}} + r^2\frac{\p\ln{\rhog}}{\p z}
  \frac{\p\ln{\tepsilon}}{\p z} \right)>0,   \\
&-\frac{\p\ln{\tepsilon}}{\p z} \notag\\
  &- \frac{\tepsilon
  h_\mathrm{g}^2}{\left(1+\tepsilon\right)^2}\frac{\Omega^2}{\kappa^2}
\frac{\p\ln{\tepsilon}}{\p\ln{r}}\left( \frac{\p\ln{\rhog}}{\p\ln{r}}\frac{\p\ln{\tepsilon}}{\p
    z} -\frac{\p\ln{\rhog}}{\p z}\frac{\p\ln{\tepsilon}}{\p\ln{
    r}}\right) > 0,
%  \kappa^2\left|\frac{\p\epsilon}{\p z}\right| + c_s^2\left|\frac{\p
%    \epsilon}{\p r}\frac{\p\epsilon}{\p z}\frac{\p\ln{\rho}}{\p
%    r}\right| - c_s^2 \left(\frac{\p\epsilon}{\p
%    r}\right)^2\left|\frac{\p\ln{\rho}}{\p z}\right| > 0.  \label{criterion2}
\end{align}
as the stability criteria, where 
\begin{align}
  h_\mathrm{g}\equiv c_s/\left(r\Omega\right)
\end{align}
is the characteristic aspect-ratio of the gas component.

%This equation shows that the vertical gradient in dust-fraction (and
%thus the dust-to-gas ratio $\rhod/\rhog$) has a stabilizing effect for
%axisymmetric perturbations. Instead, the only way to destabilize the
%system is a sufficiently large radial gradient in the dust-fraction
%and/or vertical (total) density gradient (the last term).

%Note that Eq. \ref{criterion2} is only applicable for the assume disk
%structure described above: decreasing dust and gas densities, as well
%as the dust-fraction, away from the star and away from the
%midplane. For less typical disk structures, one must evalulate above the
%critera seperately.

\subsubsection{Order-of-magnitude estimates}
We evaluate the above stability criteria by assuming the gas and dust
distributions have a Gaussian 
distribution vertically, 
\begin{align}
  \rho_\mathrm{g,d}(r,z) = \rho_\mathrm{g0,d0}(r) \exp
      {\left[-\frac{z^2}{2H^2_\mathrm{g,d}(r)}\right]}.
\end{align}
We note that $H_\mathrm{g}=h_gr$ by definition. The dust-to-gas ratio
is  
\begin{align}
  \tepsilon = \tepsilon_0(r)
  \exp{\left[-\frac{z^2}{2\widetilde{H}^2(r)}\right]}, 
\end{align}
with $\tepsilon_0= \rho_\mathrm{d,0}/\rho_{g_0}$, and  
\begin{align}
  \frac{1}{\widetilde{H}^2} \equiv \frac{1}{H_\mathrm{d}^2} -
  \frac{1}{H_\mathrm{g}^2}. 
\end{align}
We further assume the disk is approximately Keplerian so that
\begin{align}
  \kappa \simeq \Omega. 
\end{align}
%Inserting the above approximations into Eq. \ref{criterion2} and writing the result in terms of the
%dust-to-gas ratio $\tepsilon$,           
%we find 
%\begin{align}\label{ineq_est}
%  1 + \frac{\tepsilon h_\mathrm{g}^2}{\left(1 + \tepsilon\right)^2}
%  \left|\frac{\p\ln{\tepsilon}}{\p\ln{r}}\right| 
%  \left[
%    \left|\frac{\p\ln{\rhog}}{\p\ln{r}}\right| - \frac{\widetilde{H}^2}{H^2_\mathrm{g}}\left|\frac{\p\ln{\tepsilon}}{\p\ln{r}}\right|  
%    \right]>0
%\end{align}
%is required \emph{for stability.} 
%This inequality is easily satisifed in typical disk models where 
%the dust layer is much thinner than the gas scale-height, so
%that $\widetilde{H}\sim H_\mathrm{d}\ll H_\mathrm{g}$, and where the radial profiles smoothly vary on a scale of $r$.
%Then the magnitude the only destabilizing term (the third term) is negligible. According to Eq. \ref{ineq_est}, instability is on%ly possible if there is a rapid variation the radial profile of the dust-to-gas ratio.   
Then the above criteria becomes
\begin{align}
  &1 + \frac{\tepsilon}{\left(1+
    \tepsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  \frac{\p\ln{\tepsilon}}{\p\ln{r}} + 
  \frac{z^2}{\widetilde{H}^2}\right)>0 \label{stability_est1},   \\ 
&1 - \frac{\tepsilon
  h_\mathrm{g}^2}{\left(1+\tepsilon\right)^2}
  \frac{\p\ln{\tepsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\widetilde{H}^2}{H_\mathrm{g}^2}\frac{\p\ln{\tepsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
\emph{for stability}.

Eq. \ref{stability_est1}---\ref{stability_est2} are easily satisfied
in typical accretion disk models where radial gradients are $O(1/r)$ and
$h_\mathrm{g},\tepsilon, \widetilde{H}/H_\mathrm{g}\ll 1$, since then the
magnitude of the second term on the left-hand-side  of either
inequality is much less than unity. Such disks are stable against
axisymmetric perturbations.

\subsubsection{Possible instability at dust edges}
Notice the left-hand-side of Eq. \ref{stability_est2} is a quadratic in
$\p_r\tepsilon$. It is possible to
violate this inequality for sufficiently large (in magnitude) radial
gradients  in the dust-to-gas ratio, 
\begin{align}
  \frac{\p\ln{\tepsilon}}{\p\ln{r}} > S_+ \quad \mathrm{or} \quad 
  \frac{\p\ln{\tepsilon}}{\p\ln{r}} < S_-,
\end{align}
\emph{for instability}, where
\begin{align}\label{spm}
S_\pm = \frac{1}{2}\frac{H_\mathrm{g}^2}{\widetilde{H}^2} 
  \left[
  \frac{\p\ln{\rhog}}{\p\ln{r}} \pm 
  \sqrt{
  \left(\frac{\p\ln{\rhog}}{\p\ln{r}}\right)^2 + 
  4 \frac{\widetilde{H}^2}{H_\mathrm{g}^2}
  \frac{\left(1+\tepsilon\right)^2}{\tepsilon h_\mathrm{g}^2}
  }
  \,\right]. 
\end{align}

Now, if $\p_r\rhog<0$, as typical in
accretion models, then it is easier to achieve instability at a given
radius for rapidly increasing dust-to-gas ratios ($\p_r\tepsilon >
0$), and vice versa. 
%This might occur, for example, when dust is
%inserted in a finite radial range within a wider gaseous disk. 
%Physically, one might expect
%$\p_r\rhog\p_r\tepsilon > 0$ because

\emph{If} the $\p_r\rhog$ can be neglected in Eq. \ref{spm},
then instability requires
\begin{align}
  \left|\frac{\p\ln{\tepsilon}}{\p r}\right| \gtrsim
  \frac{1}{\widetilde{H}}\frac{\left(1+\tepsilon\right)}{\sqrt{\tepsilon}}. 
\end{align}
That is, if the dust-to-gas ratio varies on a lengthscale less than
$O(\widetilde{H}\sqrt{\tepsilon})$ for $\tepsilon\ll 1$, or less than
$O(\widetilde{H}/\sqrt{\tepsilon})$ for $\tepsilon\gg 1$, then the system is
potentially unstable.

\subsection{Locally isothermal gas perfectly coupled to dust} 
When $c_s(r,z)$ is non-uniform but $\tstop=0$, simple stability
criteria are not available. However, we may obtain an integral
relation for linear, axisymmetric waves with frequency $\sigma$ (see
Appendix \ref{var_prin}, where we also account for finite $\tstop$) in
the form:   
\begin{align}
&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \dd v_z \dd v_r^* B + \rho \dd v_z^*\dd v_r C +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\left|\nabla\cdot\left(P\dd
  \bm{v}\right)\right|^2\right]dV - \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV.
\end{align}
The real coefficients $A,B,C,D$ can be read off
Eq. \ref{integral_ex}. Note that $B=C$ from the assume equilibrium
structure (Eq. \ref{vshear}).

If $c_s$ is constant, then $\sigma^2$ is real, and 
the first integral leads to the usual Solberg-Hoiland criteria for
axisymmetric stability. However, when $c_s$ is non-uniform, $\sigma^2$
is generally complex, implying instability. This is responsible for
the usual vertical shear instability in locally isothermal disks.

>>>>

\begin{align}
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi - \frac{1}{\rho}\nabla
  P, \label{momeq}\\ 
  &\frac{D\epsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\epsilon 
  \tstop \nabla P \right),\label{dusteq}  
\end{align}
where $D/Dt \equiv \p_t + \bm{v}\cdot\nabla$ is the Lagragian
derivative; 
\begin{align}
  \rho \equiv \rhog + \rhod
\end{align}
is the total density with $\rhog$ and $\rhod$ being the gas and dust
density, respectively; 
\begin{align}
  \bm{v} \equiv \frac{\rhog\bm{v}_\mathrm{g} + 
    \rhod\bm{v}_\mathrm{d}}{\rho}
\end{align}
is the center-of-mass velocity with $\bm{v}_\mathrm{g}$ and
$\bm{v}_\mathrm{d}$ being the gas and dust velocities, respectively; 
and 
\begin{align}
  \epsilon \equiv \frac{\rhod}{\rho}  = \frac{\tepsilon}{1+\tepsilon} 
\end{align}
is the dust-fraction and $\tepsilon=\rhod/\rhog$ is the usual
dust-to-gas ratio. The gas pressure $P$ is given by the chosen 
equation of state described below. The external potential $\Phi$ is
that for a star of mass $M_*$ at the origin 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right), 
\end{align}
where $G$ is the gravitational constant. The second equality is the 
thin-disk approximation for the disk potential, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain analytic expressions for the disk equilibria.

@@

\begin{align}
\frac{DP}{Dt} &= - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2}
                + \mathcal{C},  \label{eff_energy} \\
\mathcal{C}&\equiv c_s^2 \nabla\cdot\left[\tstop\left(1 - \frac{P}{c_s^2\rho}\right)\nabla
  P\right].
\end{align}
Eq. \ref{eff_energy} is the energy equation for an ideal gas of adiabatic index
$\gamma=1$, i.e. isothermal evolution, but now with additional source 
terms due to the imposed temperature profile, as well as dust-gas
friction.

@@

\section{Equilibrium}\label{eqm}
To obtain an axisymmetric steady state with $\rho(r,z)$ and 
$\bm{v}=r\Omega(r,z)\hat{\bm{\phi}}$ where $\Omega = v_\phi/r$, 
we need to solve 
\begin{align}
  r\Omega^2 &= \frac{\p \Phi}{\p r} + \frac{1}{\rho}\frac{\p P}{\p
    r},\label{steady_momr}\\
  0 & = \frac{\p\Phi}{\p z} + \frac{1}{\rho}\frac{\p P}{\p z},\label{steady_momz}\\
  0 & = \nabla\cdot\left(\epsilon\tstop\nabla P\right) \label{steady_dust}
\end{align}
for $\rho$, $\Omega$ and $\epsilon$ with $P=P(\epsilon,\rho)$ given by
the equation of state (Eq. \ref{eos}).

@@

\subsection{Disk structure with a prescribed dust distribution} 
We assume a Gaussian profile in the dust-to-gas ratio, 
\begin{align}\label{dust_gauss}
  \tepsilon(r,z) = \tepsilon_0
  \exp{\left(-\frac{z^2}{2\widetilde{H}^2}\right)}. 
\end{align}
For simplicity we assume the dust-to-gas ratio at the mid-plane
$\tepsilon_0$, as well as its characteristic scale-height
$\widetilde{H}$, are both constant.

% We take the mid-plane dust-to-gas ratio
%to be a power-law in radius,  
%\begin{align}
%  \tepsilon_0(r) = \tepsilon_{00}\left(\frac{r}{r_0}\right)^{-d},  
%\end{align}
%where $\tepsilon_{00}$ is the dust-to-gas ratio at the fiducial radius
%$r_0$.

Inserting Eq. \ref{dust_gauss} into vertical hydrostatic equilibrium,
Eq. \ref{steady_momz} and integrating with the approximate
gravitational potential (Eq. \ref{thin_disk_potential}) we obtain the
gas density as
\begin{align}
  \rhog(r,z) = \rho_\mathrm{g0}(r)\exp{\left\{ - \frac{z^2}{2\Hgas^2}
    -\tepsilon_0\frac{\Htilde^2}{\Hgas^2}\left[1 -
      \exp{\left(-\frac{z^2}{2\Htilde^2}\right)}\right] \right\}}, 
\end{align}
where
\begin{align}
  \Hgas = \frac{c_s}{\OmK}, \quad \OmK \equiv \sqrt{\frac{GM_*}{r^3}},   
\end{align}
is the gas scale-height in the dust-free limit and $\OmK$ is the
Keplerian frequency, respectively.

In practice, we consider $\tepsilon_0 \ll 1$ and settled dust such that
$\Htilde<\Hgas$. Then $\rhog(r,z)$ is effectively Gaussian, as in the
dust-free case. The dust density is approximately 
\begin{align}
  \rhod \simeq \tepsilon_0\rho_\mathrm{g0}(r) \exp
        {\left(-\frac{z}{2H_\mathrm{d}^2}\right)}, 
\end{align}
with 
\begin{align}
  \frac{1}{H_\mathrm{d}^2} = \frac{1}{\Htilde^2} + \frac{1}{\Hgas^2}, 
\end{align}
and $H_\mathrm{d}$ is the dust-scale height. In numerical
calculations, we  we specify $H_\mathrm{d}< \Hgas$ to obtain 
$\Htilde$ for input.

\subsection{Orbital frequency} 
In the thin-disk approximation, the disk orbital frequency is 
\begin{align}
  \Omega(r,z) = \OmK(r)\left[1 - \frac{3}{2}\frac{z^2}{r^2} +
    \frac{h_\mathrm{g}^2}{\left(1+\tepsilon\right)}\frac{\p}{\p\ln{r}}\ln{\left(c_s^2\rhog\right)}
    \right]^{1/2}, 
\end{align}
where 
\begin{align}
  h_\mathrm{g} \equiv \frac{\Hgas}{r}
\end{align}
is the characteristic disk aspect-ratio.

\subsection{Vertical shear}
The mixture possess vertical shear. To see this, we eliminate $\Phi$
between Eq. \ref{steady_momr}---\ref{steady_momz} to 
obtain 
\begin{align}\label{vshear}
  r\frac{\p \Omega^2}{\p z} 
%&= \frac{\p\ln{\rho}}{\p r}\frac{\p}{\p
%    z}\left[c_s^2(1-\epsilon)\right] - \frac{\p\ln{\rho}}{\p z}
%  \frac{\p}{\p r} \left[c_s^2(1-\epsilon)\right]\\  
   = \frac{1}{\rho}\left(\frac{\p P}{\p r}\frac{\p s}{\p z} -\frac{\p
    P}{\p z}\frac{\p s}{\p r} \right),
\end{align}
where
\begin{align}
   s \equiv \ln \frac{P}{\rho} = \ln{\left[c_s^2(1-\epsilon)\right]} 
\end{align}
is a pseudo-entropy for the mixture.

@@

\begin{align}
  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0,\\
  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, 
\end{align}
with $s = \ln{P/\rho^\gamma}$ and 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)$ is the
square of the epicylic frequency. We assume $\kappa^2>0$. 
These criteria are applicable to any adiabatic fluid. For 
isothermal gas perfectly coupled to dust, we may take 
$s=\ln{(1-\epsilon)}$. 
%from the definition of $s$ and  
%the equation of state (Eq. \ref{eos}) with uniform sound-speed. 
%\begin{align}
%  s \equiv \ln \frac{P}{\rho} = \ln{(1-\epsilon)} 
%\end{align}
% The first equality is
%similar to the usual defintion for gas entropy (with unit adiabatic
%index). The second follows from the imposed equation of state,
%Eq. \ref{eos}, with a constant sound-speed. 
The stability conditions are thus
\begin{align}
  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\epsilon &> 0,\label{dusty_solberg1}  \\
  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\epsilon}{\p
    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \epsilon}{\p r} \right) & > 0. \label{dusty_solberg2}
\end{align}

%In accretion disks, $\rhog$, $\rhod$ and $\epsilon$ are expected to
%decrease away from the mid-plane
%\begin{align*}  
%  \frac{\p\rhog}{\p |z|} < 0,
%, \quad \frac{\p \rhog}{\p r} < 0,   
%\end{align*}
%and similarly for $\rhod$, $\epsilon$, and the dust-to-gas ratio $\tepsilon$. 
%We may thus evaluate the stability criteria at $z>0$ without loss of
%generality.

%We insert the vertical shear profile,
%Eq. \ref{vshear}, into Eq. \ref{dusty_solberg} and write the result 
%in terms of $\tepsilon$, to find

\subsubsection{Order-of-magnitude estimates} 
We evaluate the above stability criteria for the vertically-Gaussian
distributions for the gas density and the dust-to-gas ratio as 
described in \S\ref{eqm}. However, here we permit 
$\tepsilon_0=\tepsilon_0(r)$ and $\Htilde=\Htilde(r)$ to vary with
radius. We assume the disk is approximately Keplerian so that
$\kappa\simeq\OmK$. In terms of the dust-to-gas
ratio $\tepsilon$, Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2}
become   
\begin{align}
  &1 + \frac{\tepsilon}{\left(1+
    \tepsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  \frac{\p\ln{\tepsilon}}{\p\ln{r}} + 
  \frac{z^2}{\widetilde{H}^2}\right)>0 \label{stability_est1},   \\ 
&1 - \frac{\tepsilon
  h_\mathrm{g}^2}{\left(1+\tepsilon\right)^2}
  \frac{\p\ln{\tepsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\widetilde{H}^2}{H_\mathrm{g}^2}\frac{\p\ln{\tepsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
\emph{for stability}.

Eq. \ref{stability_est1}---\ref{stability_est2} are easily satisfied
in thin accretion disks where radial gradients are $O(1/r)$ and
$h_\mathrm{g},\tepsilon, \widetilde{H}/H_\mathrm{g}\ll 1$, since then the
magnitude of the second term on the left-hand-side  of either
inequality is much less than unity. Thus radially smooth, strictly
isothermal disks perfectly coupled with dust are stable against
axisymmetric perturbations.

%Such disks are stable against
%axisymmetric perturbations.

\subsubsection{Possible instability at dust edges}
Notice the left-hand-side of Eq. \ref{stability_est2} is a quadratic in
$\p_r\tepsilon$. It is possible to
violate this inequality for sufficiently large (in magnitude) radial
gradients  in the dust-to-gas ratio, 
\begin{align}
  \frac{\p\ln{\tepsilon}}{\p\ln{r}} > S_+ \quad \mathrm{or} \quad 
  \frac{\p\ln{\tepsilon}}{\p\ln{r}} < S_-,
\end{align}
\emph{for instability}, where
\begin{align}\label{spm}
S_\pm = \frac{1}{2}\frac{H_\mathrm{g}^2}{\widetilde{H}^2} 
  \left[
  \frac{\p\ln{\rhog}}{\p\ln{r}} \pm 
  \sqrt{
  \left(\frac{\p\ln{\rhog}}{\p\ln{r}}\right)^2 + 
  4 \frac{\widetilde{H}^2}{H_\mathrm{g}^2}
  \frac{\left(1+\tepsilon\right)^2}{\tepsilon h_\mathrm{g}^2}
  }
  \,\right]. 
\end{align} 
In typical accretion disks where $\p_r\rhog<0$, it is easier to
achieve instability at a given radius for increasing dust-to-gas
ratios in outwards ($\p_r\tepsilon > 0$), and vice versa.

If the $\p_r\rhog$ can be neglected in Eq. \ref{spm}, then instability
requires 
\begin{align}
  \left|\frac{\p\ln{\tepsilon}}{\p r}\right| \gtrsim
  \frac{1}{\widetilde{H}}\frac{\left(1+\tepsilon\right)}{\sqrt{\tepsilon}}. 
\end{align}
That is, if the dust-to-gas ratio varies on a lengthscale less than
$O(\widetilde{H}\sqrt{\tepsilon})$ for $\tepsilon\ll 1$, or less than
$O(\widetilde{H}/\sqrt{\tepsilon})$ for $\tepsilon\gg 1$, then the system is
potentially unstable.

\subsection{Locally isothermal gas perfectly coupled to dust} 
When $c_s(r,z)$ is non-uniform but $\tstop=0$, simple stability
criteria are not available. However, we may obtain an integral
relation for linear, axisymmetric waves with frequency $\sigma$ (see
Appendix \ref{var_prin}, where we also account for finite $\tstop$) in
the form:   
\begin{align}
&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \dd v_z \dd v_r^* B + \rho \dd v_z^*\dd v_r C +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\left|\nabla\cdot\left(P\dd
  \bm{v}\right)\right|^2\right]dV - \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV.
\end{align}
The real coefficients $A,B,C,D$ can be read off
Eq. \ref{integral_ex}. Note that $B=C$ from dynamical equilibrium
(Eq. \ref{vshear}).

If $c_s$ is constant, then $\sigma^2$ is real, and 
the first integral leads to the previous Solberg-Hoiland criteria for 
axisymmetric stability. However, when $c_s$ is non-uniform, $\sigma^2$ 
is generally complex, implying instability. This is responsible for 
the usual vertical shear instability in locally isothermal disks.
